---
title: "semibart documentation"
author: "Bret Zeldow"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{semibart documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This document is intended to provide documentation for the semibart package in R. Let $Y$ be the outcome. Currently we support $Y$ to be continuous or binary. Let $X$ be possible predictors of $Y$. Linear regression typically assumes (for continuous outcomes) that $Y = X \beta + \epsilon$ and imposes linearity among all of the $X$. Often there is just one covariate of interest--an exposure or a treatment. Call this exposure $A$ where $X = (L, A)$ and $L$ contains the remaining covariates. Since there may be interactions and nonlinearities among the relationship between $Y$ and $X$, we might not care to impose the linearity assumptions implicit in linear regression to variables $L$ that we don't care about.

Our Bayesian regression models allows some of the predictors to be modeled nonparametric with Bayesian Additive Regression Trees (BART) and other predictors to be modeled linearly. Thus, we can fit the model $Y  = f(L) + A \psi + \epsilon$, where $f(\cdot)$ is fit using BART. We can also put more terms (and interactions) in the linear part by fitting something like $Y = f(L) + A \psi_1 + A L_1 \psi_2 + \epsilon$ or $Y = f(L_{-1}) + A \psi_1 + A L_1 \psi_2 + L_1 \psi_3 + \epsilon$ where $L_{-1}$ indicates all variables in $L$ except for $L_1$.

In addition to the typical prior assumptions and distributions on the BART part (see the BayesTree package), we also specify a multivariate normal prior distribution for $\psi$. That is, $\psi \sim N(0,\sigma^2_{\psi} I).$

Our package also admits binary outcomes through the probit or logit links.


## Function and Arguments

The primary function in the semibart package is "semibart."

* x.train -- This is a matrix with each column containing a variable that is to be included in the BART part of the model
* a.train -- This is a design matrix for the linear part of the model. If there is only a single variable can use cbind(varname). If you want to include multiple variables or interactions, use cbind(varname1,varname2) or cbind(varname1,varname2,varname1*varname2). Must specifically put the interactions as a column of the matrix.
* y.train -- Vector of outcomes. Can be binary or numeric. If binary everything must be either 0 or 1. The function will automatically detect whether the outcome is binary or continuous.
* sigest -- Intial guess of the standard deviation of the regression error. If NA (the default), the function will use an estimate from least squares. This is set to 1 automatically if outcome is binary.

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
